\chapter{Literature}\doublelabel{literature}

Benveniste A., Caspi P., Edwards S.A., Halbwachs N., Le Guernic P., and
Simone R. (2003): \textbf{The Synchronous Languages Twelve Years Later}.
Proc. of the IEEE, Vol., 91, No. 1.
\href{http://www.irisa.fr/distribcom/benveniste/pub/synch_ProcIEEE_2002.pdf}{www.irisa.fr/distribcom/-benveniste/pub/synch\_ProcIEEE\_2002.pdf}

Colaco J.-L., and Pouzet M. (2003): \textbf{Clocks as First Class
Abstract Types}. In Third International Conference on Embedded Software
(EMSOFT'03), Philadelphia, Pennsylvania, USA, October 2003.\\
\url{http://www.di.ens.fr/~pouzet/lucid-synchrone/papers/emsoft03.ps.gz}

Elmqvist H., Otter M. and Cellier F.E. (1995): \textbf{Inline
Integration: A New Mixed Symbolic/Numeric Approach for Solving
Differential-Algebraic Equation Systems}. Keynote Address, Proceedings
ESM'95, European Simulation Multiconference, Prague, Czech Republic,
June 5-8, 1995, pp. xxiii-xxxiv.
% A terrible hack for long links!
\ifpdf
\href{http://citeseerx.ist.psu.edu/viewdoc/download;jsessionid=6E666F4221CFED902DCA7BDF8DC51AB6?doi=10.1.1.127.3787\&rep=rep1\&type=pdf}
{http://citeseerx.ist.psu.edu/viewdoc/download;}

\href{http://citeseerx.ist.psu.edu/viewdoc/download;jsessionid=6E666F4221CFED902DCA7BDF8DC51AB6?doi=10.1.1.127.3787\&rep=rep1\&type=pdf}
{jsessionid=6E666F4221CFED902DCA7BDF8DC51AB6?doi=10.1.1.127.3787\&rep=rep1\&type=pdf}
\else
\url{http://citeseerx.ist.psu.edu/viewdoc/download;jsessionid=6E666F4221CFED902DCA7BDF8DC51AB6?doi=10.1.1.127.3787\&rep=rep1\&type=pdf}
\fi

Forget J., F. Boniol, D. Lesens, C. Pagetti (2008): \textbf{A
Multi-Periodic Synchronous Data-Flow Language}. In
11\textsuperscript{th} IEEE High Assurance Systems Engineering Symposium
(HASE'08), Dec. 3-5 2008, Nanjing, China, pp. 251-260.
\url{http://ieeexplore.ieee.org/xpl/articleDetails.jsp?reload=true\&arnumber=4708883\&contentType=Conference+Publications}

Harel, D. (1987): \textbf{Statecharts: A Visual Formalism for Complex
Systems}\emph{.} Science of Computer Programming 8, 231-274. Department
of Applied Mathematics, The Weizmann Institute of Science, Rehovot,
Israel.
\href{http://www.inf.ed.ac.uk/teaching/courses/seoc1/2005_2006/resources/statecharts.pdf}{www.inf.ed.ac.uk/teaching/courses/seoc1/-2005\_2006/resources/statecharts.pdf}

Looye G., Thümmel M., Kurze M., Otter M., and Bals J. (2005):
\textbf{Nonlinear Inverse Models for Control}. Proceedings of
4\textsuperscript{th} International Modelica Conference, ed. G. Schmitz,
Hamburg, March 7-8.\\
\url{https://www.modelica.org/events/Conference2005/online_proceedings/Session3/Session3c3.pdf}

Pouzet M. (2006): \textbf{Lucid Synchrone, Version 3.0, Tutorial and
Reference Manual}.\\
\url{http://www.di.ens.fr/~pouzet/lucid-synchrone/}
