\chapter{Literature}\doublelabel{literature}

Benveniste A., Caspi P., Edwards S.A., Halbwachs N., Le Guernic P., and Simone R. (2003):
\bibitemtitle{The Synchronous Languages Twelve Years Later}.
Proc.\ of the IEEE, Vol., 91, No. 1.
\url{https://doi.org/10.1109/JPROC.2002.805826}

Colaco J.-L., and Pouzet M. (2003):
\bibitemtitle{Clocks as First Class Abstract Types}.
In Third International Conference on Embedded Software (EMSOFT'03), Philadelphia, Pennsylvania, USA, October 2003.\\
\url{http://www.di.ens.fr/~pouzet/lucid-synchrone/papers/emsoft03.ps.gz}

Elmqvist H., Otter M. and Cellier F.E. (1995):
\bibitemtitle{Inline Integration: A New Mixed Symbolic/Numeric Approach for Solving Differential-Algebraic Equation Systems}.
Keynote Address, Proceedings ESM'95, European Simulation Multiconference, Prague, Czech Republic, June 5-8, 1995, pp.\ xxiii-xxxiv.
\url{http://citeseerx.ist.psu.edu/viewdoc/download?doi=10.1.1.127.3787\&rep=rep1\&type=pdf}

Forget J., F. Boniol, D. Lesens, C. Pagetti (2008):
\bibitemtitle{A Multi-Periodic Synchronous Data-Flow Language}.
In 11\textsuperscript{th} IEEE High Assurance Systems Engineering Symposium (HASE'08), Dec. 3-5 2008, Nanjing, China, pp. 251-260.
\url{https://doi.org/10.1109/HASE.2008.47}

Harel, D. (1987):
\bibitemtitle{Statecharts: A Visual Formalism for Complex Systems}.
Science of Computer Programming 8, 231-274. Department of Applied Mathematics, The Weizmann Institute of Science, Rehovot, Israel.
\href{http://www.inf.ed.ac.uk/teaching/courses/seoc1/2005_2006/resources/statecharts.pdf}{www.inf.ed.ac.uk/teaching/courses/seoc1/-2005\_2006/resources/statecharts.pdf}

Looye G., Thümmel M., Kurze M., Otter M., and Bals J. (2005):
\bibitemtitle{Nonlinear Inverse Models for Control}.
Proceedings of 4\textsuperscript{th} International Modelica Conference, ed. G. Schmitz, Hamburg, March 7-8.\\
\url{https://www.modelica.org/events/Conference2005/online_proceedings/Session3/Session3c3.pdf}

Pouzet M. (2006):
\bibitemtitle{Lucid Synchrone, Version 3.0, Tutorial and Reference Manual}.\\
\url{http://www.di.ens.fr/~pouzet/lucid-synchrone/}
